---
title: Probability Distributions
output:
  blogdown::html_page:
    toc: true
author: ''
date: '2025-05-15'
math: true
toc: true
slug: probability
categories: []
tags: []
subtitle: 'A gentle introduction'
summary: 'A gentle introduction to probability'
authors: []
lastmod: '2025-05-15T11:25:37+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

<style>
body, .container, .content, .post-content, .article-content {
  max-width: 1000px !important;
  width: 100% !important;
  margin-left: auto !important;
  margin-right: auto !important;
  padding-left: 1rem;
  padding-right: 1rem;
}
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Binomial Distribution

### Generate random sequence of bernoulli trials (e.g. coin flips)
```{r rbinom function}



rbinom(n=10, # 10 times
       size=1, # one trial (i.e. one throw)
       prob=0.3) # probability of success


rbinom(n=10, # 10 times
       size=100, # 100 trials (i.e. 100 throws) - output is then #number of successes
       prob=0.3) # probability of success


```


```{r binomial distribution histogram}

flips <- rbinom(n=100000, # 10 times
          size=10, #trials (i.e. n throws) - output is then #number of successes
          prob=0.5) # probability of success

hist(x = flips)


```



### probability mass function

The probability Mass function of the binomial distribution is given by: 

$$
P(X) = \frac{n!}{x!(n-x)!} p^x q^{n-x}
$$
Where:

- \( n \): number of trials (e.g., flips)  
- \( p \): probability of success  
- \( q = 1 - p \)  
- Factorial: \( m! \) follows \( 0! = 1, 1! = 1, 2! = 2 \times 1, 3! = 3 \times 2 \times 1 \), etc.


The first part of the function is called **The binomial coefficient**, and it counts the number of ways x subjects can be drawn (choosen) from a population, n, and is expressed as n chooses x: 

$$
\binom{n}{x} = \frac{n!}{x!(n-x)!}
$$
Where: 

- \( n \): is the population 
- \( x \): is the number drawn  

```{r binom prob of x successes}

#define parameters
x <- 5 # number of success
n <- 10 # size of population 
p <- 0.5 # probability of success

#Exact probability using the probability mass function

#define  function
binom_pmf <- function(x, n, p) {
  
  q <- 1-p
  
  binom_coef <- choose(n = n, k = x)
  
  binom_coef * p^x * q^(n-x)
  
}

binom_pmf(x = x, n = n, p = p)


#Exact probability using in-built r function
dbinom(x = x, 
       size = n, 
       prob = p)

#Simulated probability
mean(rbinom(n=100000,
        size=10,
        prob=0.5) == 5)

```


## Calculate probability of at least x number of successes


```{r binom prob of at least x successes}

#Exact probability
#Cumulative function - to find at least use the complementary probability
1- pbinom(q = 4, # 4 or less, (or at least five if using the complementary) 
       size = 10, # number of throws
       prob = 0.5) # probability

#or use the lower.tail = FALSE (P[X > x])
pbinom(q = 4, # 4 or less, (or at least five if using the complementary) 
       size = 10, # number of throws
       prob = 0.5, lower.tail = FALSE) # probability


#Simulated probability
mean(rbinom(n=100000,
        size=10,
        prob=0.5) >= 5)



#simulate several probabilities with different size using map
n <- c(100, 1000, 10000, 100000)


map_dbl(.x = n, ~mean(rbinom(n = .x,
        size=10,
        prob=0.5) >= 5))



```
## Expected value and variance



```{r binom expectation and variance}

#Expected value
size <- 100
prob <- 0.8

#Simulation
mean(rbinom(n=10, # 10 times
       size=size, # 100 trials (i.e. 100 throws) - output is then #number of successes
       prob=prob)) # probability of success

#Expected value rule
size*prob


#variance

#simulation
var(rbinom(n=10, # 10 times
       size=size, # 100 trials (i.e. 100 throws) - output is then #number of successes
       prob=prob)) # probability of success

#Variance rule
size*prob*(1-prob)


```


```{r sda}
density <- function(x) {20/x^2}

integrate(density, lower = 10, upper = 20)



dbinom(2, 5, 0.9)


```