---
title: In Situ Acceleration-Speed Profile
author: 'Lasse Ishøi'
date: '2023-03-13'
categories: []
tags: []
subtitle: "Automisation using using the CatapultR package"
summary: "A walkthrough of how to calculate the acceleration-speed profile in R"
authors: []
lastmod: '2024-01-22T13:18:49+01:00'
featured: yes
image: 
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

The In Situ Acceleration-Speed Profile has gained increasing popularity since introduced by JB Morin and team in 2021. Briefly described, the Acceleration-Speed profile is in-game mapping of the maximal acceleration and speed capabilities based on instantaneous acceleration and speed data from GPS or equivalent tracking from multiple training sessions/games. For an in depth overview of the concept, a good place to start is the __[blogpost](https://jbmorin.net/2021/07/29/the-in-situ-sprint-profile-for-team-sports-testing-players-without-testing-them/)__ by JB Morin and the following scientific papers: 

* Paper 1
* Paper 2
* Paper 3

This post serves as a walkthrough of how the calculation can be made relatively easy in R without no need to download raw data csv files manually. The prerequisite is Catapult GPS data and the __[catapultR package](http://catapultr.catapultsports.com/)__ which serves as a wrapper for the catapult API. 

I start of by creating a group of functions based on the catapultR package. I load the following packages:
```{r load libraries, echo=TRUE, message=FALSE}

library(catapultR)
library(tidyverse)
library(lubridate)



```

and set a few placeholder variables and the API connection:
```{r placeholder, echo=TRUE, eval=FALSE}

#Set player name, and date range for the data of interest
playerName <- name
startDate <- startdate
endDate <- enddate


#Get API Access
sToken <- stoken
sRegion <- sregion


#Setup API connection
token <- ofCloudCreateToken(sToken = sToken,  sRegion = sRegion)

```

The next code chunk uses several functions from the catapultR package to retrieve the raw 10HZ sensor data via the API connection. We use the three parameters __playerName__, __startDate__, and __endDate__ to specify the data of interest.

```{r get data from API, echo=TRUE, eval=FALSE}

#overview of athletes connected to the Token user
athletes <- ofCloudGetAthletes(token)

#Use playerName to identify associated athlete id
athletes$player_name <- paste(athletes$first_name, athletes$last_name)
athleteID <- athletes$id[athletes$player_name==playerName]

# get activies from the specified period
from <- as.integer(as.POSIXct(as.Date(startDate, format="%d-%m-%Y")))
to <- as.integer(as.POSIXct(as.Date(endDate, format="%d-%m-%Y")))
activities <- ofCloudGetActivities(token, from = from, to = to)

#loop over the activities for the specific athlete to get the 10Hz raw data
#Create output variable as a list using the vector function with length equal to number of activities
rawData <- vector("list", nrow(activities))


#for loop using the ofCloudGetActivitySensorData function which takes the athleteID and an activity as arguments. 
#"Try" function suppresses an error in the for loop if the athelte is not part of an activity.
for (i in seq_along(activities)) {try(
rawData[[i]] <- ofCloudGetActivitySensorData(
  token,
  athlete_id = athleteID,
  activity_id = activities$id[[i]],
  parameters = c("ts", 
                 "cs", 
                 "lat", 
                 "long", 
                 "xy", 
                 "o", 
                 "v", 
                 "rv", 
                 "a", 
                 "hr", 
                 "pl", 
                 "sl", 
                 "mp", 
                 "pq", 
                 "ref", 
                 "hdop")), 
  silent = TRUE)
}

#map function to unnest the data (gps variables) column across the list. 
#"Possibly" function secures that the unnest function does not stop due to error

rawData_list <- rawData %>%
                            map(possibly(~unnest(data = ., cols = "data"), NULL)) %>%
#delete empty lists - these are activites where the player of interest has not participated
                            compact()

#mutate a time column using luridate based on ts (epoch), and rename columns
rawData_list <-  rawData_list %>% map(~mutate(.data = ., timestamp = as_datetime(.$ts),
                                                          seconds = round(cs/100, 1),
                                                          elapsed_time = (row_number()-1)/10)) %>%
                                  map(~rename(.data = ., latitude = lat,
                                              longitude = long,
                                              position_x = x,
                                              position_y = y,
                                              velocity = v,
                                              velocity_raw = rv,
                                              acceleration = a,
                                              total_distance = o,
                                              heart_rate = hr,
                                              player_load = pl,
                                              smoothed_load = sl,
                                              metabolic_power = mp,
                                              positional_quality = pq,
                                              number_satelites = ref))


```

To make it easier going forward I wrap the above code in a function:

```{r get data function,  echo=TRUE, eval=FALSE}

get_raw_period_data <- function(tag_name, filter_period, startdate, enddate, stoken, sregion) {

  playerName <- name
startDate <- startdate
endDate <- enddate


sToken <- stoken
sRegion <- sregion

token <- ofCloudCreateToken(sToken = sToken,  sRegion = sRegion)

athletes <- ofCloudGetAthletes(token)

athletes$player_name <- paste(athletes$first_name, athletes$last_name)
athleteID <- athletes$id[athletes$player_name==playerName]

from <- as.integer(as.POSIXct(as.Date(startDate, format="%d-%m-%Y")))
to <- as.integer(as.POSIXct(as.Date(endDate, format="%d-%m-%Y")))
activities <- ofCloudGetActivities(token, from = from, to = to)

rawData <- vector("list", nrow(activities))


for (i in seq_along(activities)) {try(
rawData[[i]] <- ofCloudGetActivitySensorData(
  token,
  athlete_id = athleteID,
  activity_id = activities$id[[i]],
  parameters = c("ts", 
                 "cs", 
                 "lat", 
                 "long", 
                 "xy", 
                 "o", 
                 "v", 
                 "rv", 
                 "a", 
                 "hr", 
                 "pl", 
                 "sl", 
                 "mp", 
                 "pq", 
                 "ref", 
                 "hdop")), 
  silent = TRUE)
}


rawData_list <- rawData %>%
                map(possibly(~unnest(data = ., cols = "data"), NULL)) %>%
                compact()

rawData_list <-  rawData_list %>% map(~mutate(.data = ., timestamp = as_datetime(.$ts),
                                 seconds = round(cs/100, 1),
                                 elapsed_time = (row_number()-1)/10)) %>%
                                  map(~rename(.data = ., latitude = lat,
                                              longitude = long,
                                              position_x = x,
                                              position_y = y,
                                              velocity = v,
                                              velocity_raw = rv,
                                              acceleration = a,
                                              total_distance = o,
                                              heart_rate = hr,
                                              player_load = pl,
                                              smoothed_load = sl,
                                              metabolic_power = mp,
                                              positional_quality = pq,
                                              number_satelites = ref))
}
```

The final step I need to solve before I can use the function is to identify the __sToken__ and the __sRegion__ (These are text strings used to access your user profile via the API, and can can be found in the Catapult Cloud). With that in place, I am ready to collect some raw 10 Hz data by simply adding the parameters to the function:

```{r get data example,  echo=TRUE, eval=FALSE}

#I want to get data from Player A and in the data range 5th to 10th January 2024
data <- get_raw_period_data(name = "Player A", 
                            startdate = "05-01-2024", 
                            enddate = "10-01-2024", 
                            stoken = "xxx", 
                            sregion = "xxx")

```

The data comes in a list format, with each list representing an activity. In this case, I get 4 activities (training session), which may be on the low end for obtaining a reliable profile, but that is not the main focus for now.

```{r load the data, echo=FALSE}

#Preview of the data
load("data.RData")



```

```{r preview data}

#Preview of the data (selecting a few variables)
data[[1]] %>% 
  select(athlete_first_name, velocity, velocity_raw, acceleration, elapsed_time) %>%
  top_n(10)


```


```{r AS profile}

library(signal)
library(runner)
library(useful)
library(InSituASProfile)

#Set filter details
bf4 <- butter(2, 0.15, type="low") 


#Apply filter
rawData_list <- data %>%
                    #Delete NA's
                    map(~na.omit(.)) %>%
                    map(~mutate(.data = ., velocity_bw = filtfilt(bf4, .$velocity))) %>%
                    map(~mutate(.data = ., velocity_raw_bw = filtfilt(bf4, .$velocity_raw)))



rawData_list_acc <- rawData_list %>% 
                    map(~mutate(.data = ., 
                                reg_acc = runner(x = ., 
                                                 k = 10, 
                                                 na_pad = FALSE, 
                                                 f = function(x){
                                                 model <- lm(velocity_raw_bw ~ elapsed_time,
                                                 data = x)
                                  coefficients(model)[2]})))


rawData_list_shifted <- rawData_list_acc %>%
                                     map(~shift.column(data = ., 
                                                       columns = "reg_acc",
                                                       newNames = "acc", 
                                                       up = TRUE, len = 2))


all_data <- rawData_list_shifted %>%
                            bind_rows()

###derive the acceleration from speed and time

 all_data <- rawData_list %>%
                             bind_rows()
  all_data$accel_bw <- c(NA, with(all_data, diff(velocity_bw)/diff(elapsed_time)))
  all_data$acc <- c(NA, with(all_data, diff(velocity_raw_bw)/diff(elapsed_time)))
  
  ##########################################



trial_mutate_slice <- all_data %>% slice(40476:40550)

trial_mutate_slice <- trial_mutate_slice %>% mutate(time = row_number())

ggplot(trial_mutate_slice, aes (x = elapsed_time)) +
  geom_line(aes(y = velocity), color = "red") +
  geom_line(aes(y = velocity_bw), color = "green") +
  geom_line(aes(y= velocity_raw_bw), color = "blue") + 
  geom_line(aes(y = acceleration), color = "black") + 
  geom_line(aes(y = acc), color = "orange") +
  theme_bw()

 all_data %>%
  ggplot(aes(x = velocity_raw_bw, y = acc)) + 
        geom_point(size = 0.2) + 
        ylim(0,NA) + 
        geom_vline(xintercept = 3, color = "red")
 
 
 all_data <- all_data %>% 
            rename(speed = velocity_raw_bw)

InSituASProfile::prepare_data(all_data,print_plot = TRUE)
InSituASProfile::get_AS_Profile()
  

```

